An algorithm for the optimization of finite element integration loops

### 2016-12-30 TOMS_Paper/reviewers_response.txt

First of all, we would like to thank the referees for their insightful comments,
which helped improving the paper. In this new version, we have attempted to
address all of the referees' concerns. We explain how the paper has changed
below, by considering each of the requests/observations/questions raised by the
three referees individually.

@@

| My suggestions are as follows:
| 1. It would be better to include actual transformed code and original code in
| the appendix for all computing model (mass, helmholtz, elasticity, hyper
| elasticity). At least I wonder what makes hugh speed up in the hyper elastic
| problem compared to usual mass matrix.

The generated code has been uploaded to GitHub and given a DOI through Zenodo.
This is properly documented in the paper (see Section 7.1 - Experimental setup).
For reasons of space, it simply was not possible to append code in an appendix.

| 2. Please include actual FLOP measurements of input code and transformed code
| and show some correlation with performance.

There is a whole new section treating this fundamental aspect (7.8 - On the
correlation between operation count and performance). Further, a new figure was
added (Fig. 11): this clearly shows the relationship between the achieved
speed-ups and FLOPs reductions. We believe that this figure is particularly
useful, because it shows how well the system presented in the paper scales well
with the complexity of the operators. It also provides a nice summary of the
performance improvements achieved. Moreover, Sections 7.4 through 7.7 (Mass,
Helmholtz, ...) have been extended and now also discuss the impact of the reduced
operation counts in a set of relevant test cases.

@@

Referee: 2
==========

| The paper is well written, interesting does clearly merit publication.
| The usage of advanced compiler technology as demonstrated here is clearly
| novel and it brings accelerated performance.

| There is one issue that is not covered in the analysis and only briefly
| mentioned in the general discussion: the transfer from memory. I understand
| that it complicates the analysis but the least it should be quantified in
| the numerical experiments [...] It is to a certain extent discussed already
| in the paper but it should also be quantified.

Sections 7.4 through 7.7 now discuss the memory transfer cost, but only in a
limited set of test cases. We believe that it is very difficult to thoroughly
study this problem in a paper like ours, being based on a high dimensional
experimentation space. A complete treatment would probably require a new paper,
introducing specific metrics and showing different angles of the performance
achieved.  However, there exists a very recent paper including exactly the type
of analysis advocated by the referee: [Bercea et al.  2016]. This is now
referenced in Section 7.2, along with a high level discussion of the memory
transfer impact.  It is worth noting that the Bercea paper is based upon the same
Firedrake technology of our article.

### 2017-01-13 TOMS_Paper/reviewers_response.txt

Referee: 1
==========

| This paper is overall well written. Examples are very illustrative and help
| understand the article. My only concern is about how authors relates their
| theortical results with numerical performance examples. [...]

| My suggestions are as follows:
| 1. It would be better to include actual transformed code and original code in
| the appendix for all computing model (mass, helmholtz, elasticity, hyper
| elasticity). At least I wonder what makes hugh speed up in the hyper elastic
| problem compared to usual mass matrix.

The generated code has been uploaded to GitHub and permanently
archived with a DOI through Zenodo.  This is properly documented in
the paper (see Section 7.1 - Experimental setup).  For reasons of
space, it simply was not possible to append code in an appendix.

@@

Referee: 2
==========

| The paper is well written, interesting does clearly merit publication.
| The usage of advanced compiler technology as demonstrated here is clearly
| novel and it brings accelerated performance.

| There is one issue that is not covered in the analysis and only briefly
| mentioned in the general discussion: the transfer from memory. I understand
| that it complicates the analysis but the least it should be quantified in
| the numerical experiments [...] It is to a certain extent discussed already
| in the paper but it should also be quantified.

Sections 7.4 through 7.7 now discuss the memory transfer cost, but
only in a limited set of test cases. We believe that it is very
difficult to thoroughly study this problem in a paper like ours, being
based on a high dimensional experimentation space. A complete
treatment would probably require a new paper, introducing specific
metrics and showing different angles of the performance achieved.
However, with other coauthors we have very recently published a paper
including exactly the type of analysis advocated by the referee:
[Bercea et al.  2016]. This is now referenced in Section 7.2, along
with a high level discussion of the memory transfer impact.  It is
worth noting that the Bercea paper is based upon the same Firedrake
technology of our article.

@@

| 3) Results: The authors present several charts with speedups on different
| problems, which gives detailed information. However, it would be nice to add a
| figure with aggregate results, such as a scatter plot of all instances or a
| performance profile.

We agree with the referee about the usefulness of such a chart. We have added a
scatter plot (Fig. 11) as well as a section explaining what information we can
extrapolate from it.

| 4) The manuscript is already quite long so if possible, do not increase the
| total length (i.e. try find things to tighten or cut).

We tried our best, but we feel that we cannot really stay within 27
pages without losing essential information. We observe, however, that
this paper is within the typical range of page lengths in TOMS.

### 2017-01-16 TOMS_Paper/assembly.tex

We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots. The axes represent $q$, $\mathrm{nf}$, and code transformation system. We show one subplot for each problem instance ${\langle} \textrm{form}, \mathrm{nf}, q {\rangle}$, with the code transformation system varying within each subplot. The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over the code produced by the FEniCS Form Compiler with no transformations applied (also referred to as ``baseline``). We note that if a bar or a subplot are missing, then the form compiler failed to generate code because it either exceeded the system memory limit or was otherwise unable to handle the form.

The rest of the section is organized as follows: we first provide insights into the general outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we motivate, for each form, the performance results obtained and the relationship with the operation count.

\subsection{High level view}
Our transformation strategy does not always guarantee minimum execution time. In particular, about 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4,\ \mathrm{nf}=0]$. There are two reasons for this. First, low level optimization can have a significant impact on the actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Second, the transformation strategy adopted when $T_H$ is exceeded plays a key role, as we will later elaborate.

The memory transfer has sometimes a non-negligible impact, especially when low order polynomials are used. This is more evident in the simpler bilinear forms (e.g., \texttt{Mass}), when the speed-ups over the baseline are limited. The effect of the memory transfer is quantified, to some extent, in the sections below. For a comprehensive analysis of its cost, the reader is invited to refer to~\cite{extrusion-paper}, a study based upon the same Firedrake technology used in this article.

\subsection{Autovectorization}
We chose the mesh dimension and the function spaces such that the inner loop sizes would always be a multiple of the machine vector length. This ensured autovectorization in the majority of code variants\footnote{We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}. The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization \cite{SLP-vect} is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.

@@

We first notice the lack of improvements when $q=1$. The operation count of \texttt{cfO2} and \texttt{tensor} is roughly 9$\times$ lower than that of the alternatives (tens instead of hundreds operations), but nonetheless no gain in execution time is obtained. This is due to the fact that local assembly is memory-bound. For instance, in the simplest case $nf=0$, in addition to the determinant of the Jacobian inverse, \texttt{tens} only requires three unique multiplications to evaluate the local element matrix (26 operations in total), whereas almost a hundred memory accesses are performed. In particular, we have: 12 loads for the vertex coordinates; for each vertex, 1 load to indirectly access its data; for each of the 16 local elements, 4 loads (optimistically) and 1 store to update an entry in the global matrix in compressed sparse row format. This unsurprisingly leads to a very low arithmetic intensity (i.e., $\frac{\text{operation count}}{\text{bytes moved}}$), which collocates the kernels in a memory-bound regime.

@@

A relevant experiment we additionally performed was relaxing the memory threshold by setting $T_H = \operatorname{size}(\mathrm{L3})$. We found that this makes \texttt{cfO2} generally slower for $\mathrm{nf} \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle \mathrm{nf}=2, q=2\rangle$. This effect could be worse when running in parallel, since the L3 cache is shared and different threads would end up competing for the same resource.

@@

We also tried to relate the achieved performance to the operation counts obtained by each transformation system. We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups were generally massive. In the original code, $37 \cdot 10^9$ operations were counted at the source level. However, the number of operations actually executed is in practice much lower, since by inspection of the assembly code generated by the underlying compiler it is clear that common sub-expressions elimination and loop-invariant code motion are applied in aggressive way. The actual operation count is, however, definitely larger than the $10^8$ operations performed by \texttt{ufls}. \texttt{cfO2} brings the operation count down to $3 \cdot 10^7$; that is, more than 3$\times$ lower than \texttt{ufls}. This is completely in line with the drop in execution times, with \texttt{cfO2} outperforming \texttt{ufls} by slightly more than 3$\times$.

>>>>

We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots. The axes represent $q$, $\mathrm{nf}$, and code transformation system. We show one subplot for each problem instance ${\langle} \textrm{form}, \mathrm{nf}, q {\rangle}$, with the code transformation system varying within each subplot. The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over the code produced by the FEniCS Form Compiler with no transformations applied (also referred to as ``baseline``). We note that if a bar or a subplot is missing, then the form compiler failed to generate code because it either exceeded the system memory limit or was otherwise unable to handle the form.

The rest of the section is organized as follows: we first provide insights into the general outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we discuss, for each form, the performance results obtained and the relationship with the operation count.

\subsection{High level view}
Our transformation strategy does not always guarantee minimum execution
time. In particular, about 5$\%$ of the test cases (3 out of 56, excluding
marginal differences) \texttt{cfO2} did not produce the code with the lowest
runtime. The most significant of such test cases is the elastic model with $[q=4,\ \mathrm{nf}=0]$. There are two reasons for this. First, low level optimization can have a significant impact on the actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Second, the transformation strategy adopted when $T_H$ is exceeded plays a key role, as we will later elaborate.

The cost of transferring data from memory sometimes has a noticable impact
on performance, especially when
low order polynomials are used. This is more evident in the simpler bilinear
forms (e.g., \texttt{Mass}), for which the speed-ups over the baseline are
limited. The effect of memory transfer costs is quantified, to some extent, in the sections below. For a comprehensive analysis of its cost, the reader is referred to~\cite{extrusion-paper}, a study based upon the same Firedrake technology used in this article.

\subsection{Autovectorization}
We chose the mesh dimension and the function spaces such that the inner loop
sizes are always a multiple of the machine vector length. This ensures
autovectorization in the majority of code variants\footnote{We verified the
  vectorization of inner loops by looking at both compiler reports and
  assembly code.}. The most prominent exception is \texttt{quad}, where the
presence of indirection arrays in the generated code impedes
vectorization. In \texttt{tens}, loop nests are fully unrolled, so the
standard loop vectorization is not feasible; the compiler reports suggest,
however, that block vectorization \cite{SLP-vect} is often triggered. In
\texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have
identical structure, and the compuler regularly applies loop vectorization.

@@

We first notice the lack of improvements when $q=1$. The operation count of
\texttt{cfO2} and \texttt{tensor} is roughly 9$\times$ lower than that of
the alternatives (tens instead of hundreds of operations), but nonetheless no gain in execution time is obtained. This is due to the fact that local assembly is memory-bound. For instance, in the simplest case $nf=0$, in addition to the determinant of the Jacobian inverse, \texttt{tens} only requires three unique multiplications to evaluate the local element matrix (26 operations in total), whereas almost a hundred memory accesses are performed. In particular, we have: 12 loads for the vertex coordinates; for each vertex, 1 load to indirectly access its data; for each of the 16 local elements, 4 loads (optimistically) and 1 store to update an entry in the global matrix in compressed sparse row format. This unsurprisingly leads to a very low arithmetic intensity (i.e., $\frac{\text{operation count}}{\text{bytes moved}}$), which locates the kernels in a memory-bound regime.

@@

A relevant experiment we additionally performed was relaxing the memory
threshold by setting $T_H = \operatorname{size}(\mathrm{L3})$. We found that
this makes \texttt{cfO2} generally slower for $\mathrm{nf} \geq 2$, with a
maximum slow-down of 2.16$\times$ with $\langle \mathrm{nf}=2,
q=2\rangle$. This effect might be expected to become worse when running in parallel, since the L3 cache is shared and different threads would end up competing for the same resource.

@@

It is also possible to relate the achieved performance to the operation counts
obtained by each transformation system. We here limit ourselves to consider
the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups were
generally very large. In the original code, $37 \cdot 10^9$ operations were counted at the source level. However, the number of operations actually executed is in practice much lower, since by inspection of the assembly code generated by the underlying compiler it is clear that common sub-expressions elimination and loop-invariant code motion are applied in aggressive way. The actual operation count is, however, definitely larger than the $10^8$ operations performed by \texttt{ufls}. \texttt{cfO2} brings the operation count down to $3 \cdot 10^7$; that is, more than 3$\times$ lower than \texttt{ufls}. This is completely in line with the drop in execution times, with \texttt{cfO2} outperforming \texttt{ufls} by slightly more than 3$\times$.
